Smooth and Persistent Forecasts of German GDP

### 2025-08-08 Archive IWH Simon paper/23.06.2025/paper.tex
up

\begin{equation}
    u_{t+h} = \beta' \cdot \mathbf{X}_t + \varepsilon_t
    \label{eq:direct_forecast}
\end{equation}

where
\begin{itemize}
    \item $u_{t+h}$ is the growth rate of GDP, $h = \{0, 1, 2, ... \}$
    \item $\mathbf{X}_t \equiv [1,u_{t-1}, ip_t, ifo\_c_t, ESI_t, spr\_10y\_3m_t]'$ or their transformed counterparts and $u_{t-1}$ accounts for the publication lag of GDP.\footnote{Including lagged GDP enables the nesting of the AR forecast within the broader modeling framework.}
    \item $\hat{\beta}$ is estimated by least-squares, omitting observations affected by the pandemic, i.e. Q4-2019 to Q4-2020.
\end{itemize}

@@

Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} compare actual and fitted values for GDP at forecast horizons $h = 0, \ldots, 3$, using the four indicators and the target, i.e. GDP shifted forward (leftwards) by $h$ quarters, without applying any filtering. Figure \ref{direct_wc} shows results over the sample without the pandemic, while Figure \ref{direct_wc_financial_crisis} provides a closer look at the fit during the financial crisis. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients on the indicators are zero (HAC-adjusted Wald test).

@@

Figures \ref{direct_hp_forecasts} and \ref{direct_hp_forecasts_financial_crisis} provide results comparable to those of Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} but now using indicators smoothed with the HP-C filter, covering the full sample without the pandemic and providing a closer look at performance during the financial crisis.  The fitted values lag GDP slightly less and track turning points somewhat more quickly. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients $\hat{\beta} = 0$. The use of filtered indicators improves their statistical significance at forecast horizons $h > 2$ quarters. 
%Figures \ref{direct_hp_forecasts} (all observations with exclusion of the pandemic) and \ref{direct_hp_forecasts_financial_crisis} (financial crisis). A comparison with the classic (unfiltered) direct forecasts in the latter figure %in Fig.\eqref{direct_wc_financial_crisis} 
% indicates that the new predictors are slightly less retarded and track the turning points more promptly (`faster'). 
% The filtered approach also exhibits superior statistical properties. As shown in Table \ref{tab:f_stats}, 
%The p-values of the F-statistics for the HP-based forecasts are smaller than for the direct forecasts specifically at $h=3$, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators).

@@

%\begin{table}[ht]
%\centering
%\begin{tabular}{rrrrr}
%  \hline
% & h=3 & h=4 & h=5 & h=6 \\ 
%  \hline
%rRMSE & 0.962 & 0.967 & 0.990 & 0.967 \\ 
%   \hline
%\end{tabular}
%\caption{Relative root mean-square error of HP-C against classic direct forecast for $h\geq %3$.\\Values $<1$ imply  HP-C has smaller RMSE.  } 
%\label{tab:rRMSE}
%\end{table}
\begin{table}[h]
\centering
\begin{tabular}{lrrrrrr}
  \hline
  & h= 1 & h= 2 & h= 3 & h= 4 & h= 5 & h= 6 \\ 
  \hline
  Unfiltered & 0.000 & 0.001 & 0.910 & 0.509 & 0.091 & 0.178 \\ 
HP-C filtered & 0.001 & 0.036 & 0.046 & 0.049 & 0.012 & 0.022 \\ 
   \hline
\end{tabular}
\caption{p-values for $H_0: \hat{\beta} = 0$, HAC-adjusted Wald test.\\Estimation over full sample (without pandemic.)} 
\label{tab:pvaluedhp}
\end{table}

@@

Eq. \eqref{critssaht} has several important implications. First, the objective function is indifferent to an affine transformation of the predictor. This ambiguity can be resolved by assuming an arbitrary scale and level for $y_t$ (standardization). Alternatively, a mean square error norm (MSE) can be substituted for the target correlation, as noted in \cite{Wildi2025}. In this case, the classic MSE predictor $y_{t,MSE}$ is obtained as a solution to the SSA criterion by insertion of $1/HT_1:=1/HT_{MSE}$ in the constraint, where $HT_{MSE}$ represents the holding time of $y_{t,MSE}$. %If the MSE predictor is noisy, which is often the case in applications, one can select $HT_1>HT_{MSE}$  in the constraint to address the rate of noisy crossings or false alarms. 
Second, the concept can be extended to a multivariate framework, denoted M-SSA, by generalizing the correlation functions as detailed in \cite{Wildi2025}. 
%Lastly, sign changes in trend growth, represented by zero-crossings of the target $z_t$, are indicative of relevant changes in the trajectory of the economy, for example, at transitions between expansions and recessions. 
Lastly, when $z_t$ is GDP growth, sign changes are indicative of expansions and recessions.
Unfortunately, classic predictors often generate excessively many false (`noisy') sign changes due to high-frequency leakage, as discussed above.
%as illustrated by the amplitude function in Figure \eqref{hp_160} (bottom right panel) and further elaborated upon by \cite{Wildi2025}. 
 \cite{Wildi2025} shows that the SSA criterion helps control this phenomenon,  see Fig.\ref{mssa_msse_zc} for illustration. Using the equivalent dual formulation of the criterion \eqref{critssaht}, the author demonstrates that the SSA solution exhibits the lowest zero-crossing rate of any (linear) predictor with the same target correlation. In the following, we therefore use the M-SSA framework to construct a `smooth' predictor $y_t$ to forecast $z_t$. %In a second stage, serving as a generalization and refinement of the univariate HP-C utilized in the direct HP forecast.

@@

For illustration, consider the problem of nowcasting HP-filtered GDP, denoted by HP-GDP, where the target $z_t$ is generated using the two-sided HP(160) filter examined previously. We compare the classic multivariate mean squared error (MSE) predictor (denoted M-MSE) with the M-SSA predictor where we set $HT_1=1.5\times HT_{MSE}$. This constraint ensures that the M-SSA predictor produces roughly 33$\%$ fewer sign changes than M-MSE in large samples. Comparisons with M-MSE allow us to see the cost of this constraint in terms of prediction performance (target correlation). 
Both predictors use the same underlying VAR(1) model introduced in Section \ref{sec:direct_forecast} and estimated on data up to Q4-2007. This ensures a lengthy out-of-sample period that includes significant events such as the financial crisis, the sovereign debt crisis, and the COVID-19 pandemic\footnote{In the parsimonious VAR(1) model estimated over the shorter sub-sample ending in Q4-2007, the economic sentiment indicator (ESI) does not receive any weight. Additional technical details and background information are provided in \cite{Wildi2025}. All computations are conducted using the M-SSA package, as documented in \cite{Wildi2025}. The package can be accessed at \url{https://github.com/wiaidp/R-package-SSA-Predictor}.}

@@

\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=4.5in]{./Figures/mssa_msse_zc.pdf}\caption{Sign changes of the M-MSE (green) and M-SSA (blue) nowcasts towards the financial crisis, with sign changes of the M-MSE marked by green vertical lines. The sample holding time (HT) of each filter corresponds to the average duration between its consecutive zero-crossings. For each filter, its sample HT converges to the respective expected HT asymptotically. Here, we impose  $HT_{MSSA}=1.5\times HT_{MSE}$: in the long run, we expect the M-MSE to exhibit approximately $50\%$ more sign changes than the M-SSA.\label{mssa_msse_zc}}\end{center}\end{figure}
% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Wed May  7 12:49:47 2025
\begin{table}[ht]
\centering
\begin{tabular}{lrr}
  \hline
 & M-SSA & M-MSE \\ 
  \hline
Target correlation & 0.68 & 0.69 \\ 
  HT & 6.50 & 3.71 \\ 
   \hline
\end{tabular}
\caption{Sample target correlations and holding time (HT) of M-SSA and M-MSE nowcasts.} 
\label{corhtnow}
\end{table}

Figure \ref{bk_gammak} displays the coefficients of the multivariate nowcasts, where M-SSA's increased smoothing reduces the maximum amplitudes of the filter weights and slows their rate of decay at longer lags. The resulting filtered series, presented in Figure \ref{mssa_msse_now}, are standardized to enable direct comparisons. The M-SSA nowcast is slightly smoother than that of the M-MSE, exhibiting approximately one-third fewer sign-changes. Fig.\ref{mssa_msse_zc} compares both nowcasts in relation to the financial crisis, highlighting that the classic (MSE-based) design tends to produce unwanted `noisy' zero-crossings at (or in the vicinity of) the transitions between recessions and expansions. Further, Table \ref{corhtnow} compares sample target correlations and HTs of the two nowcasts. Increased smoothness, i.e. holding time increases from 3.7 to 6.5 quarters, achieved by the M-SSA comes at the cost of a slight reduction (0.68 vs 0.69) in target correlation. This highlights the inherent trade-off—often referred to as the prediction dilemma—between smoothness (zero-crossing rate) and predictive accuracy (target correlation) within the M-SSA criterion. 
The results validate that the smoothing constraint incorporated into the multivariate extension of criterion \eqref{critssa} effectively manages the sign-changes of the predictor: sample estimates of the holding times align with the expected values derived from Eq.~\eqref{ht}.\footnote{Effective convergence of sample estimates to their theoretical expected values —assuming the true data-generating process is known— can be validated using the M-SSA package. This validation is typically performed through analyses of very long samples of simulated (multivariate) data, ensuring the consistency of the estimation procedures under idealized conditions.} An increase of $50\%$ in the HT relative to the M-MSE predictor has only a modest effect on the target correlation, suggesting that this trade-off could be beneficial in our application. %Going forward, we now omit the M-MSE, as it is a special case of the more general M-SSA framework.

@@

The statistical significance of the (out of sample) GDP component predictor in Equation \ref{oos_pred}  can be evaluated by regressing 
$\textrm{GDP}_{t+shift}$ on $\hat{\textrm{GDP}}_{t+shift}(h)$:
\begin{equation}\label{eq_gdp_pred}
\textrm{GDP}_{t+shift}=\alpha_0+\alpha_1\hat{\textrm{GDP}}_{t+shift}(h)+\nu_{t}^{shift,h},
\end{equation}
for observations after Q4-2007: the significance of the corresponding HAC-adjusted statistics is indicated by a `star' symbol in Table \ref{rRMSE_mSSA_comp_mean_wc}\footnote{A single star means significance at the $5\%$ level; a double star means significance at the $0.1\%$ level. Note that statistical significance is influenced by the relatively short out-of-sample span, of length 17 years.}. The MSE forecast performance of $\hat{\textrm{GDP}}_{t+shift}(h)$ can be benchmarked against the mean predictor (see Table \ref{rRMSE_mSSA_comp_mean_wc}) as well as against the direct forecast (see Table \ref{rRMSE_mSSA_comp_direct_without_covid7}). For consistency, both benchmark predictors are also quarterly updated out-of-sample designs.\\

@@

Our results can be summarized as follows. When excluding the pandemic, the direct forecasts outperform the mean forecast up to two quarters ahead, see Table \ref{rRMSE_mSSA_direct_mean_without_covid8} in the appendix. %\footnote{However, the pandemic period generally obscures the relationships between predictors and GDP.}.
In contrast, the GDP component predictors $\hat{\textrm{GDP}}_{t+shift}(h)$, optimized for longer forecast horizons $h\geq 5$, outperform the mean benchmark up to 5Q ahead, see Table \ref{rRMSE_mSSA_comp_mean_wc}. Additionally,  $\hat{\textrm{GDP}}_{t+shift}(h)$, optimized for large forecast horizons $h\geq 4$, surpass the direct forecasts for $shift\geq 3$, see Table \ref{rRMSE_mSSA_comp_direct_without_covid7} in the appendix. Statistical significance of the GDP component predictor, optimized for larger forecast horizons, can be established for at least up to one year ahead, see Table \ref{rRMSE_mSSA_comp_mean_wc}. Finally, the strong performance of predictors optimized for larger forecast horizons ($h\geq 4$), at smaller shifts ($shifts< 4$) corroborates the importance of the left-shift of the corresponding predictors in relation to forecast performance.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 0.975 & 0.941^{*} & 0.886^{*} & 0.839^{**} & 0.843^{**} & 0.897^{**} & 0.965 \\ 
  Shift=1 & 1.085 & 1.068 & 1.011 & 0.927^{*} & 0.871^{**} & 0.875^{**} & 0.919^{**} \\ 
  Shift=2 & 1.076 & 1.089 & 1.074 & 1.011 & 0.938^{*} & 0.902^{**} & 0.909^{**} \\ 
  Shift=3 & 1.037 & 1.069 & 1.090 & 1.059 & 0.992 & 0.943^* & 0.930^* \\ 
  Shift=4 & 1.013 & 1.058 & 1.101 & 1.097 & 1.037 & 0.967 & 0.925^* \\ 
  Shift=5 & 1.016 & 1.062 & 1.106 & 1.115 & 1.078 & 1.015 & 0.955^* \\  
   \hline
\end{tabular}
\caption{rRMSEs: out-of-sample forecast errors Equation \ref{oosfe} vs expanding mean of GDP\\Out-of-sample evaluation starting in Q1-2008, without pandemic.\\Values $< 1$ indicate superior predictions from Equation \ref{oos_pred}\\
Row-index: Shift refers to the left-shift of the target $\textrm{GDP}_{t+shift}$\\
Column-index: $h$ refers to the forecast horizon for which $\textrm{M-SSA}_t^{\textrm{GDP}}(h)$ is optimized\\
Significance refers to the hypothesis $\alpha_1=0$ in Equation \ref{eq_gdp_pred}.
} 
\label{rRMSE_mSSA_comp_mean_wc}
\end{table}

@@

% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Wed May  7 12:49:51 2025
% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Wed May  7 12:49:51 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 1.128 & 1.086 & 1.030 & 0.985 & 0.992 & 1.044 & 1.109 \\ 
  Shift=1 & 1.280 & 1.254 & 1.189 & 1.096 & 1.037 & 1.041 & 1.081 \\ 
  Shift=2 & 1.173 & 1.181 & 1.166 & 1.104 & 1.031 & 0.995 & 0.995 \\ 
  Shift=3 & 1.012 & 1.037 & 1.058 & 1.040 & 0.989 & 0.949 & 0.934 \\ 
  Shift=4 & 0.982 & 1.016 & 1.056 & 1.063 & 1.022 & 0.974 & 0.941 \\ 
  Shift=5 & 0.961 & 0.997 & 1.041 & 1.058 & 1.028 & 0.979 & 0.936 \\ 
   \hline
\end{tabular}
\caption{Regression of shifted GDP on M-SSA($h$) GDP component.\\rRMSEs benchmarked against the direct forecast.\\Out-of-sample evaluation starting in Q1-2008, without pandemic.} 
\label{rRMSE_mSSA_comp_direct_without_covid7}
\end{table}% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Wed May  7 12:49:51 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrr}
  \hline
 & Shift=0 & Shift=1 & Shift=2 & Shift=3 & Shift=4 & Shift=5 \\ 
  \hline
rRMSE & 0.862 & 0.838 & 0.912 & 1.008 & 1.001 & 1.028 \\  
   \hline
\end{tabular}
\caption{Direct forecasts.\\
rRMSEs benchmarked against the expanding mean.\\Out-of-sample evaluation starting in Q1-2008, without pandemic.} 
\label{rRMSE_mSSA_direct_mean_without_covid8}
\end{table}

>>>>

\begin{equation}
    u_{t+h} = \boldsymbol{\beta'} \cdot \mathbf{X}_t + \varepsilon_t
    \label{eq:direct_forecast}
\end{equation}

where
\begin{itemize}
    \item $u_{t+h}$ is the growth rate of GDP, $h = \{0, 1, 2, ... \}$
    \item $\mathbf{X}_t \equiv [1,u_{t-1}, ip_t, ifo\_c_t, ESI_t, spr\_10y\_3m_t]'$ or their transformed counterparts and $u_{t-1}$ accounts for the publication lag of GDP.\footnote{Including lagged GDP enables the nesting of the AR forecast within the broader modeling framework.}
    \item $\hat{\boldsymbol{\beta}}$ is estimated by least-squares, omitting observations affected by the pandemic, i.e. Q4-2019 to Q4-2020.
\end{itemize}

@@

Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} compare actual and fitted values for GDP at forecast horizons $h = 0, \ldots, 3$, using the four indicators and the target, i.e., GDP shifted leftwards by $h$ quarters, without applying any filtering. Figure \ref{direct_wc} shows results over the sample without the pandemic, while Figure \ref{direct_wc_financial_crisis} provides a closer look at the fit during the financial crisis. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients on the indicators are zero (HAC-adjusted Wald test).

@@

Figures \ref{direct_hp_forecasts} and \ref{direct_hp_forecasts_financial_crisis} provide results comparable to those of Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} but now using indicators smoothed with the HP-C filter, covering the full sample without the pandemic and providing a closer look at performance during the financial crisis.  The fitted values lag GDP slightly less and track turning points somewhat more quickly. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients ${\boldsymbol{\beta}} = 0$. The use of filtered indicators improves their statistical significance at forecast horizons $h > 2$ quarters. 
%Figures \ref{direct_hp_forecasts} (all observations with exclusion of the pandemic) and \ref{direct_hp_forecasts_financial_crisis} (financial crisis). A comparison with the classic (unfiltered) direct forecasts in the latter figure %in Fig.\eqref{direct_wc_financial_crisis} 
% indicates that the new predictors are slightly less retarded and track the turning points more promptly (`faster'). 
% The filtered approach also exhibits superior statistical properties. As shown in Table \ref{tab:f_stats}, 
%The p-values of the F-statistics for the HP-based forecasts are smaller than for the direct forecasts specifically at $h=3$, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators).

@@

%\begin{table}[ht]
%\centering
%\begin{tabular}{rrrrr}
%  \hline
% & h=3 & h=4 & h=5 & h=6 \\ 
%  \hline
%rRMSE & 0.962 & 0.967 & 0.990 & 0.967 \\ 
%   \hline
%\end{tabular}
%\caption{Relative root mean-square error of HP-C against classic direct forecast for $h\geq %3$.\\Values $<1$ imply  HP-C has smaller RMSE.  } 
%\label{tab:rRMSE}
%\end{table}
\begin{table}[h]
\centering
\begin{tabular}{lrrrrrr}
  \hline
  & h= 1 & h= 2 & h= 3 & h= 4 & h= 5 & h= 6 \\ 
  \hline
  Unfiltered & 0.000 & 0.001 & 0.910 & 0.509 & 0.091 & 0.178 \\ 
HP-C filtered & 0.001 & 0.036 & 0.046 & 0.049 & 0.012 & 0.022 \\ 
   \hline
\end{tabular}
\caption{p-values for $H_0: \boldsymbol{\beta} = 0$, HAC-adjusted Wald test.\\Estimation over full sample (without pandemic.)} 
\label{tab:pvaluedhp}
\end{table}

@@

Eq. \eqref{critssaht} has several important implications. First, the objective function is indifferent to an affine transformation of the predictor. This ambiguity can be resolved by assuming an arbitrary scale and level for $y_t$ (standardization). Alternatively, a mean square error norm (MSE) can be substituted for the target correlation, as noted in \cite{Wildi2025}. In this case, the classic MSE predictor $y_{t,MSE}$ is obtained as a solution to the SSA criterion by insertion of $1/HT_1:=1/HT_{MSE}$ in the constraint, where $HT_{MSE}$ represents the holding time of $y_{t,MSE}$. %If the MSE predictor is noisy, which is often the case in applications, one can select $HT_1>HT_{MSE}$  in the constraint to address the rate of noisy crossings or false alarms. 
Second, the concept can be extended to a multivariate framework, denoted M-SSA, by generalizing the correlation functions as detailed in \cite{Wildi2025}. 
%Lastly, sign changes in trend growth, represented by zero-crossings of the target $z_t$, are indicative of relevant changes in the trajectory of the economy, for example, at transitions between expansions and recessions. 
Lastly, when $z_t$ is GDP growth, sign changes are indicative of expansions and recessions.
Unfortunately, classic predictors often generate excessively many false (`noisy') sign changes due to high-frequency leakage, as discussed above.
%as illustrated by the amplitude function in Figure \eqref{hp_160} (bottom right panel) and further elaborated upon by \cite{Wildi2025}. 
 \cite{Wildi2025} shows that the SSA criterion helps control this phenomenon,  see Fig.\ref{mssa_msse_zc} for illustration. Using the dual formulation of the criterion \eqref{critssaht}, the author demonstrates that the SSA solution yields the lowest zero-crossing rate among all (linear) predictors with equivalent target correlation. Consequently, the subsequent analysis employs the M-SSA framework to construct a `smooth' predictor $y_t$ for forecasting $z_{t+h}$. In particular, it is shown that controlling the zero-crossing rate within the M-SSA framework has a quantifiable effect on the hit rate and false positive rate.

 %In a second stage, serving as a generalization and refinement of the univariate HP-C utilized in the direct HP forecast.

@@

For illustration, consider the problem of nowcasting HP-filtered GDP, denoted by HP-GDP, where the target $z_t$ is generated using the two-sided HP(160) filter examined previously. We compare the classic multivariate mean squared error (MSE) predictor (denoted M-MSE) with the M-SSA predictor where we set $HT_1=1.5\times HT_{MSE}$. This constraint ensures that the M-SSA predictor produces roughly 33$\%$ fewer sign changes than M-MSE in large samples. Comparisons with M-MSE allow us to see the cost of this constraint in terms of prediction performance (target correlation). 
Both predictors use the same underlying VAR(1) model introduced in Section \ref{sec:direct_forecast} and estimated on data up to Q4-2007. This ensures a lengthy out-of-sample period that includes significant events such as the financial crisis, the sovereign debt crisis, and the COVID-19 pandemic.\footnote{The economic sentiment indicator (ESI) does not receive any weight in the parsimonious VAR(1) model estimated over the shorter sub-sample ending in Q4-2007. Technical details and background information on the M-SSA are provided in \cite{Wildi2025}. All computations are conducted using the M-SSA package, as documented in \cite{Wildi2025}. The package can be accessed at \url{https://github.com/wiaidp/R-package-SSA-Predictor}.}

@@

\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=4.5in]{./Figures/mssa_msse_zc.pdf}\caption{Sign changes of the M-MSE (green) and M-SSA (blue) nowcasts towards the financial crisis, with sign changes of the M-MSE marked by green vertical lines. The sample holding time (HT) of each filter corresponds to the average duration between its consecutive zero-crossings. For each filter, its sample HT converges to the respective expected HT asymptotically. Here, we impose  $HT_{MSSA}=1.5\times HT_{MSE}$: in the long run, we expect the M-MSE to exhibit approximately $50\%$ more sign changes than the M-SSA.\label{mssa_msse_zc}}\end{center}\end{figure}
% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Wed May  7 12:49:47 2025
\begin{table}[ht]
\centering
\begin{tabular}{lrr}
  \hline
 & M-SSA & M-MSE \\ 
  \hline
Target correlation & 0.68 & 0.69 \\ 
  HT & 6.50 & 3.71 \\ 
   \hline
\end{tabular}
\caption{Sample target correlations and holding times (HT) of M-SSA and M-MSE nowcasts.} 
\label{corhtnow}
\end{table}

Figure \ref{bk_gammak} displays the coefficients of the multivariate nowcasts, where M-SSA's increased smoothing reduces the maximum amplitudes of the filter weights and slows their rate of decay at longer lags. The resulting filtered series, presented in Figure \ref{mssa_msse_now}, are standardized to enable direct comparisons. The M-SSA nowcast is slightly smoother than that of the M-MSE, exhibiting approximately one-third fewer sign-changes. Fig.\ref{mssa_msse_zc} compares both nowcasts in relation to the financial crisis, highlighting that the classic (MSE-based) design tends to produce unwanted `noisy' zero-crossings at (or in the vicinity of) the transitions between recessions and expansions. Further, Table \ref{corhtnow} compares sample target correlations and HTs of the two nowcasts. Increased smoothness, i.e. holding time increases from 3.7 to 6.5 quarters, achieved by the M-SSA comes at the cost of a slight reduction (0.68 vs 0.69) in target correlation. This highlights the inherent trade-off—often referred to as the prediction dilemma—between smoothness (zero-crossing rate) and predictive accuracy (target correlation) within the M-SSA criterion. 
The results validate that the smoothing constraint incorporated into the multivariate extension of criterion \eqref{critssa} effectively manages the sign-changes of the predictor: sample estimates of the holding times align with the expected values derived from Eq.~\eqref{ht}.\footnote{Effective convergence of sample estimates to their theoretical expected values —assuming the true data-generating process is known— can be validated using the M-SSA package. This validation is typically performed through analyses of very long samples of simulated (multivariate) data, ensuring the consistency of the estimation procedures under idealized conditions.} An increase of $50\%$ in the HT relative to the M-MSE predictor has only a modest effect on the target correlation, suggesting that this trade-off could be beneficial in our application, as confirmed by a subsequent analysis of hit rate and false alarm rate. %Going forward, we now omit the M-MSE, as it is a special case of the more general M-SSA framework.

@@

The statistical significance of the (out of sample) GDP component predictor in Equation \ref{oos_pred}  can be evaluated by regressing 
$\textrm{GDP}_{t+shift}$ on $\hat{\textrm{GDP}}_{t+shift}(h)$:
\begin{equation}\label{eq_gdp_pred}
\textrm{GDP}_{t+shift}=\alpha_0+\alpha_1\hat{\textrm{GDP}}_{t+shift}(h)+\nu_{t}^{shift,h},
\end{equation}
for observations after Q4-2007: the significance of the corresponding HAC-adjusted statistics is indicated by a `star' symbol in Table \ref{rRMSE_mSSA_comp_mean_wc}\footnote{A single star means significance at the $5\%$ level; a double star means significance at the $0.1\%$ level. Note that statistical significance is influenced by the relatively short out-of-sample span, of length 17 years.}. The MSE forecast performance of $\hat{\textrm{GDP}}_{t+shift}(h)$ can be benchmarked against the mean predictor (see Table \ref{rRMSE_mSSA_comp_mean_wc} which reports relative Root Mean-Square Errors, rRMSE) as well as against the direct forecast (see Table \ref{rRMSE_mSSA_comp_direct_without_covid7}). For consistency, both benchmark predictors are also quarterly updated out-of-sample designs.\\

@@

Our results can be summarized as follows. When excluding the pandemic, the direct forecasts outperform the mean forecast up to two quarters ahead, see Table \ref{rRMSE_mSSA_direct_mean_without_covid8} in the appendix. %\footnote{However, the pandemic period generally obscures the relationships between predictors and GDP.}.
In contrast, the GDP component predictors $\hat{\textrm{GDP}}_{t+shift}(h)$, optimized for longer forecast horizons $h\geq 5$, outperform the mean benchmark up to 5Q ahead, see Table \ref{rRMSE_mSSA_comp_mean_wc}. Additionally,  $\hat{\textrm{GDP}}_{t+shift}(h)$, optimized for large forecast horizons $h\geq 4$, surpass the direct forecasts for $shift\geq 3$, see Table \ref{rRMSE_mSSA_comp_direct_without_covid7} in the appendix. Statistical significance of the GDP component predictor, optimized for larger forecast horizons, can be established for at least up to one year ahead, see Table \ref{rRMSE_mSSA_comp_mean_wc}. Finally, the strong performance of predictors optimized for larger forecast horizons ($h\geq 4$), at smaller shifts ($shifts< 4$) corroborates the importance of the left-shift of the corresponding predictors, see Figs.\ref{multivar_vs_univar} and \ref{mssa_hpc_financial_crisis}, in relation to forecast performance.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 0.975 & 0.941^{*} & 0.886^{*} & 0.839^{**} & 0.843^{**} & 0.897^{**} & 0.965 \\ 
  Shift=1 & 1.085 & 1.068 & 1.011 & 0.927^{*} & 0.871^{**} & 0.875^{**} & 0.919^{**} \\ 
  Shift=2 & 1.076 & 1.089 & 1.074 & 1.011 & 0.938^{*} & 0.902^{**} & 0.909^{**} \\ 
  Shift=3 & 1.037 & 1.069 & 1.090 & 1.059 & 0.992 & 0.943^* & 0.930^* \\ 
  Shift=4 & 1.013 & 1.058 & 1.101 & 1.097 & 1.037 & 0.967 & 0.925^* \\ 
  Shift=5 & 1.016 & 1.062 & 1.106 & 1.115 & 1.078 & 1.015 & 0.955^* \\  
   \hline
\end{tabular}
\caption{rRMSEs: out-of-sample forecast errors Equation \ref{oosfe} vs expanding mean of GDP\\Out-of-sample evaluation starting in Q1-2008, without pandemic\\Values $< 1$ indicate superior predictions from Equation \ref{oos_pred}\\
Row-index: Shift refers to the left-shift of the target $\textrm{GDP}_{t+shift}$\\
Column-index: $h$ refers to the forecast horizon for which $\textrm{M-SSA}_t^{\textrm{GDP}}(h)$ is optimized\\
Significance refers to the hypothesis $\alpha_1=0$ in Equation \ref{eq_gdp_pred}.
} 
\label{rRMSE_mSSA_comp_mean_wc}
\end{table}

@@

% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Wed May  7 12:49:51 2025
% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Wed May  7 12:49:51 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 1.131 & 1.092 & 1.028 & 0.973 & 0.977 & 1.040 & 1.119 \\ 
  Shift=1 & 1.295 & 1.274 & 1.206 & 1.106 & 1.039 & 1.044 & 1.097 \\ 
  Shift=2 & 1.180 & 1.194 & 1.178 & 1.109 & 1.029 & 0.990 & 0.997 \\ 
  Shift=3 & 1.029 & 1.061 & 1.082 & 1.051 & 0.984 & 0.936 & 0.923 \\ 
  Shift=4 & 1.012 & 1.057 & 1.100 & 1.096 & 1.036 & 0.967 & 0.924 \\ 
  Shift=5 & 0.988 & 1.033 & 1.076 & 1.084 & 1.048 & 0.987 & 0.929 \\  
   \hline
\end{tabular}
\caption{Regression of shifted GDP on M-SSA($h$) GDP component.\\rRMSEs benchmarked against the direct forecast.\\Out-of-sample evaluation starting in Q1-2008, without pandemic.} 
\label{rRMSE_mSSA_comp_direct_without_covid7}
\end{table}% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Wed May  7 12:49:51 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrr}
  \hline
 & Shift=0 & Shift=1 & Shift=2 & Shift=3 & Shift=4 & Shift=5 \\ 
  \hline
rRMSE & 0.862 & 0.838 & 0.912 & 1.008 & 1.001 & 1.028 \\  
   \hline
\end{tabular}
\caption{Direct forecasts.\\
rRMSEs benchmarked against the expanding mean.\\Out-of-sample evaluation starting in Q1-2008, without pandemic.} 
\label{rRMSE_mSSA_direct_mean_without_covid8}
\end{table}
